### 2025-07-25 paper/preprint/appendix.tex

Based on Equation~\ref{eq-loglike-decomp} and Equation~\ref{eq-multi} we
can identify that
\((\mu_{k,d}-\mu_{m,d}) \propto (\theta_{k,d} - \theta_{m,d})\) under
the assumptions we made above. Hence, we have that
\((\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}}) < 0\)
and
\((\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}}) > 0\)

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-train-default}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-eval-default}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, opacityback=0]

@@

To calculate the contributions, IG compares the output to a baseline.
The selection of an appropriate baseline is an important design decision
--- it should produce a ``neutral'' prediction to avoid capturing
effects that cannot be directly attributed to the model
(\citeproc{ref-sundararajan2017ig}{Sundararajan, Taly, and Yan 2017};
\citeproc{ref-sturmfels2020visualizing}{Sturmfels, Lundberg, and Lee
2020}). To remain consistent in our evaluations, we use a baseline drawn
at random from a uniform distribution, \(\mathcal{U}(-1,1)\) all
datasets. This aligns with standard evaluation practices for IG.

@@

As IG outputs are not bounded (i.e., they are arbitrary real numbers),
it becomes a challenge to meaningfully compare IG outputs of different
models --- ones that are trained conventionally, and ones that underwent
counterfactual training. For our purposes, we observe that the direction
of contribution (sign) does not matter. If protecting a feature flips
the sign of the corresponding output, then CT „overcorrected'' for this
feature, but as long as the magnitude of its contribution decreases,
then CT achieved its objective according to our formulation. Thus, we
focus on the magnitude of IG outputs in the evaluation. Next, to
accommodate for variable ranges of outputs, we standardize the outputs.

Let \(\mathbf{g}_d\) denote the estimated IG for feature \(d\). Then in
the case of 2D synthetic datasets we find that taking the absolute value
of the outputs, \(|\mathbf{g}_d|\), and then dividing them by a
\(\max(\mathbf{g}) - \min(\mathbf{g})\) term allows us to make the most
meaningful comparison. In the case of real-world datasets we choose to
normalize the values to a 0-1 range instead.

We compare the (average) sensitivity to the features that were protected
for CT models. We again use bootstrapping (100 rounds, 2500 samples per
round) to establish the significance of our results. These are presented
in Table~\ref{tbl-ig}.

@@

\begin{minipage}{0.50\linewidth}

\begin{table}[H]

@@

\end{minipage}%
%
\begin{minipage}{0.50\linewidth}

\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/mnist_ig.png}}

\caption{\label{fig-mnist-ig}Class-conditional standardized integrated
gradients for MNIST. Top: BL. Bottom: CT. Mutability constraints were
imposed on the five top and bottom rows of pixels.}

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-train-final-run-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-tune-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-tune_lr-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, opacityback=0]

>>>>

Based on Equation~\ref{eq-loglike-decomp} and Equation~\ref{eq-multi} we
can identify that
\((\mu_{k,d}-\mu_{m,d}) \propto (\theta_{k,d} - \theta_{m,d})\) under
the assumptions we made above. Hence, we have that
\((\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}}) < 0\)
and
\((\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}}) > 0\).

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-train-default}: Training Phase}, opacityback=0, colback=white]

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-eval-default}: Evaluation Phase}, opacityback=0, colback=white]

@@

To calculate the contributions, IG compares the output to a baseline.
The selection of an appropriate baseline is an important design decision
--- it should produce a ``neutral'' prediction to avoid capturing
effects that cannot be directly attributed to the model
(\citeproc{ref-sundararajan2017ig}{Sundararajan, Taly, and Yan 2017};
\citeproc{ref-sturmfels2020visualizing}{Sturmfels, Lundberg, and Lee
2020}). To remain consistent in our evaluations, we use a baseline drawn
at random from a uniform distribution, \(\mathcal{U}(-1,1)\), for all
datasets. This aligns with standard evaluation practices for IG.

@@

As IG outputs are not bounded (i.e., they are arbitrary real numbers),
it becomes a challenge to meaningfully compare IG outputs of different
models --- ones that are trained conventionally, and ones that underwent
counterfactual training. For our purposes, we observe with reference to
our Proposition, that we are interested estimating changes in the
relative contribution of protected features compared to mutable ones.
Thus, to meaningfully compare integrated gradients for different models
and to accommodate for variable ranges of outputs in absolute terms, we
standardize the integrated gradients across features.

Let \(\mathbf{g}_d\) denote the estimated IG for feature \(d\). Then in
the case of 2D synthetic datasets we find that taking the absolute value
of the outputs, \(|\mathbf{g}_d|\), and then dividing them by a
\(\max(\mathbf{g}) - \min(\mathbf{g})\) term allows us to make the most
meaningful comparison. In the case of real-world datasets we choose to
normalize the values to a \([0,1]\) range instead. We compare the
(average) sensitivity to the features that were protected for CT models.
Once again we use bootstrapping (100 rounds, 2500 samples per round) to
establish the significance of our results. These are presented in
Table~\ref{tbl-ig}.

@@

@@

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, opacityback=0, colback=white]

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, opacityback=0, colback=white]

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, opacityback=0, colback=white]

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, opacityback=0, colback=white]

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, opacityback=0, colback=white]

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, opacityback=0, colback=white]

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-tune-train}: Training Phase}, opacityback=0, colback=white]

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, opacityback=0, colback=white]

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-tune_lr-train}: Training Phase}, opacityback=0, colback=white]

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, opacityback=0, colback=white]

### 2025-07-25 paper/preprint/paper.tex

\caption{\label{fig-mnist}Visual explanations for \emph{MNIST} for BL
(top) and CT (bottom). \textbf{Plausibility}: col.~1 is a random factual
0 (blue); cols. 2-5 are corresponding \emph{ECCCo} counterfactuals in
target classes 1 to 4. \textbf{Actionability}: cols. 6-10 show
integrated gradients averaged over test images in classes 5 to 9.}

@@

For synthetic datasets, we always protect the first feature; for all
real-world tabular datasets we could identify and protect an \emph{age}
variable; for \emph{MNIST}, we protect the five upper and lower pixel
rows of the full image. Statistically significant reductions in costs
overwhelmingly point in the expected positive direction reaching up to
around 60\% for \emph{GMSC}. Only in the case of \emph{Cred}, average
costs increase, likely because any potential benefits from protecting
the \emph{age} are outweighed by the increase in costs required for
greater plausibility. The findings for \emph{Adult} and \emph{MNIST} are
not significant. A qualitative inspection of the class-conditional
integrated gradients in Figure~\ref{fig-mnist} (columns 6-10) suggests
that CT still has the expected effect: the model (bottom) is insensitive
(blue) to the protected rows of pixels; details of this experiment are
reported in the supplementary appendix.

>>>>

\subcaption{\label{fig-mnist-ce}\emph{Plausibility}: \emph{ECCCo}
counterfactuals for factual `0' (blue).}

\end{minipage}%
\newline
\begin{minipage}{\linewidth}

\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/mnist_ig.png}}

\subcaption{\label{fig-mnist-ig}\emph{Actionability}: class-cond.
integrated gradients.}

\end{minipage}%

\caption{\label{fig-mnist}Visual explanations for \emph{MNIST}. Top and
bottom rows (of images) show the results for BL and CT, respectively.
Mutability constraints are imposed on the five top and bottom rows (of
pixels). CT produces plausible counterfactuals (a) and is less sensitive
to protected features (b).}

@@

For synthetic datasets, we always protect the first feature; for all
real-world tabular datasets we could identify and protect an \emph{age}
variable; for \emph{MNIST}, we protect the five top and bottom pixel
rows of the full image. Statistically significant reductions in costs
overwhelmingly point in the positive direction reaching up to around
60\% for \emph{GMSC}. Only in the case of \emph{Cred}, average costs
increase, likely because any benefits from protecting \emph{age} are
outweighed by an increase in costs required for greater plausibility.
The findings for \emph{Adult} and \emph{MNIST} are insignificant. A
qualitative inspection of the class-conditional integrated gradients in
Figure~\ref{fig-mnist-ig} suggests that CT still has the expected
effect: the model (bottom) is less sensitive (blue) to the protected
rows of pixels; details are reported in the appendix.

### 2025-07-25 paper/presenations/slides.qmd

---
title: counterfactual training
subtitle: Update Meeting Dec 2024
institute: Delft University of Technology
date: today
format: 
  beamer:
    theme: Boadilla
  tudelft-revealjs:
    theme: custom.scss
    self-contained: true
    smaller: false
    scrollable: true
    preview-links: auto
    slide-number: true
    transition: slide
    background-transition: fade
    fig-align: center
    html-math-method: mathjax
    include-in-header:
      - text: |
          <script>
          MathJax = {
            options: {
              menuOptions: {
                settings: {
                  assistiveMml: false
                }
              }
            }
          };
          </script>
crossref: 
  prp-title: RQ
  prp-prefix: RQ
classoption: "notheorems"
---

>>>>

---
title: counterfactual training
subtitle: Update Meeting Dec 2024
date: today
format: 
  beamer:
    theme: Boadilla
  tudelft-revealjs:
    theme: custom.scss
    self-contained: true
    smaller: false
    scrollable: true
    preview-links: auto
    slide-number: true
    transition: slide
    background-transition: fade
    fig-align: center
    html-math-method: mathjax
    include-in-header:
      - text: |
          <script>
          MathJax = {
            options: {
              menuOptions: {
                settings: {
                  assistiveMml: false
                }
              }
            }
          };
          </script>
crossref: 
  prp-title: RQ
  prp-prefix: RQ
classoption: "notheorems"
---

### 2025-07-25 paper/sections/experiments.qmd

![*Plausibility*: faithful counterfactuals (*ECCCo*).](/paper/figures/mnist_ce.png){#fig-mnist-ce}

@@

Visual explanations for *MNIST*. Top and bottom rows (of images) show the results for BL and CT, respectively. Mutability constraints are imposed on the five top and bottom rows (of pixels). CT produces plausible counterfactuals (a) and is insensitive to protected features (b).

@@

For synthetic datasets, we always protect the first feature; for all real-world tabular datasets we could identify and protect an *age* variable; for *MNIST*, we protect the five top and bottom pixel rows of the full image. Statistically significant reductions in costs overwhelmingly point in the positive direction reaching up to around 60% for *GMSC*. Only in the case of *Cred*, average costs increase, likely because any benefits from protecting *age* are outweighed by an increase in costs required for greater plausibility. The findings for *Adult* and *MNIST* are insignificant. A qualitative inspection of the class-conditional integrated gradients in @fig-mnist-ig suggests that CT still has the expected effect: the model (bottom) is insensitive (blue) to the protected rows of pixels; details are reported in the appendix.

>>>>

![*Plausibility*: *ECCCo* counterfactuals for factual '0' (blue).](/paper/figures/mnist_ce.png){#fig-mnist-ce}

@@

Visual explanations for *MNIST*. Top and bottom rows (of images) show the results for BL and CT, respectively. Mutability constraints are imposed on the five top and bottom rows (of pixels). CT produces plausible counterfactuals (a) and is less sensitive to protected features (b).

@@

For synthetic datasets, we always protect the first feature; for all real-world tabular datasets we could identify and protect an *age* variable; for *MNIST*, we protect the five top and bottom pixel rows of the full image. Statistically significant reductions in costs overwhelmingly point in the positive direction reaching up to around 60% for *GMSC*. Only in the case of *Cred*, average costs increase, likely because any benefits from protecting *age* are outweighed by an increase in costs required for greater plausibility. The findings for *Adult* and *MNIST* are insignificant. A qualitative inspection of the class-conditional integrated gradients in @fig-mnist-ig suggests that CT still has the expected effect: the model (bottom) is less sensitive (blue) to the protected rows of pixels; details are reported in the appendix.
